Fully structured proof theory for intuitionistic modal logics

### 2020-07-02 BicolorNS/aiml20/aiml20.tex
final version completed + galmiche bib + macros inside tex

\input{macros}

@@

\section{Introduction}

Structural proof theoretic accounts of modal logic can adopt the paradigm of \emph{labelled deduction} in the form of e.g.~labelled sequent systems~\cite{vigano2000,negri2005}, or the one of \emph{unlabelled deduction} in the form of e.g.~nested sequent systems~\cite{brunnler2009deep,poggiolesi2009method}.

@@

Similarly, proof systems have been designed for intuitionistic modal logic both as labelled sequent systems~\cite{simpson1994} and as nested sequent systems~\cite{strassburger2013,kuznets:strassburger:maehara}.
%
Surprisingly, in both settings, extending the sequent structure with the same \emph{one} extra element is enough to obtain sound and complete systems.

@@

Incorporating the preorder symbol into the syntax, however,  allowed us~\cite{marin:morales:strassburger:hal} to %also obtain a sound and complete proof system for the intuitionistic modal logic extended with axiom $\rn{4}_\rn\DIA$, by designing the following rule:
%$$\scalebox{.9}{$\vlinf{\diabrn{4}}{\text{\footnotesize $\lb{u'}$ fresh}}{\B, \accs wv, \accs vu, \Left \SEQ \Right}{\B, \accs wv, \accs vu, \futs{u}{u'}, \accs{w}{u'}, \Left \SEQ \Right}$}$$
%Consequently, we can easily 
translate the conditions in~\eqref{eq:4frame} into separate inference rules for $\ax{4_{\BOX}}$ and $\ax{4_{\DIA}}$:\vspace*{-.2cm}
\begin{equation*}\small
\label{eq:4rules}
\vlinf{\rn{4_{\BOX}}}{\text{\footnotesize$\lb{u}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
	\B, \accs xy, \accs yz, \accs {u}z, \futs x{u},\Left \SEQ \Right}
\quad
\vlinf{\rn{4_{\DIA}}}{\text{\footnotesize$\lb{u}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
	\B, \accs xy, \accs yz, \accs {x}{u}, \futs z{u},\Left \SEQ \Right}  
\end{equation*}
%This allows us to define cut-free deductive systems for a wide range of logics that could not be treated before.

@@

The generalisation of these results to our fully structured framework is subject of ongoing study.

>>>>

%\input{macros}
\definecolor{notgreen}{rgb}{.1,.6,.1}
\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
\newcommand{\red}[1]{{\color{red}{#1}}}
\newcommand{\blue}[1]{{\color{blue}{#1}}}
\newcommand{\marianela}[1]{{\color{purple}[Marianela: #1]}}
\newcommand{\sonia}[1]{{\color{blue}[Sonia: #1]}}

%%% General

\newcommand{\F}{\mathcal{F}}
\newcommand{\M}{\mathcal{M}}
%\newcommand{\G}{\mathcal{G}}

\newcommand{\B}{\mathcal{B}}
\newcommand{\Left}{\mathcal{L}}
\newcommand{\Right}{\mathcal{R}}

\newcommand*{\rel}{R}
\newcommand*{\rn}[1]  {\ensuremath{\mathsf{#1}}}
\newcommand*{\invr}[1]{#1^\mathsf{inv}}

\newcommand*{\orn}[2][]  {\rn{#2}_{#1}}%^{\lab}}}
\newcommand*{\rrn}[2][]  {\rn{#2}_\rn{R#1}}%^\lab}}
\newcommand*{\lrn}[2][]  {\rn{#2}_\rn{L#1}}%^\lab}}

\newcommand*{\fm}[1]{#1}%{{\color{notgreen}{#1}}}
\newcommand*{\lb}[1]{#1}%{{\color{blue}{#1}}}
\newcommand*{\labels}[2]{\lb{#1}\mathord{:}\fm{#2}}
\newcommand*{\accs}[2]{\lb{#1}R\lb{#2}}
\newcommand*{\futs}[2]{\lb{#1}\le{\lb{#2}}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Logic
\newcommand*{\ax}[1]{\mathsf{#1}}
\newcommand*{\kax}[1][]		{\ax{k_{#1}}}
\newcommand*{\A}{\mathcal{A}}

\newcommand{\force}[2]{#1\Vdash#2}
\newcommand{\nforce}[2]{#1\nVdash#2}
%
\newcommand{\height}[1]{|#1|}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Systems
\newcommand*{\K}{\mathsf{K}}
\newcommand*{\IK}{\mathsf{IK}}
\newcommand*{\lab}{\mathsf{lab}}
\newcommand*{\n}{\mathsf{n}}

%%% Connectives
\newcommand*{\NOT}{\neg}
\newcommand*{\AND}{\mathbin{\wedge}}
\newcommand*{\TOP}{\mathord{\top}}
\newcommand*{\OR}{\mathbin{\vee}}
\newcommand*{\BOT}{\mathord{\bot}}
\newcommand*{\IMP}{\mathbin{\scalebox{.9}{\raise.2ex\hbox{$\supset$}}}}

\newcommand*{\BOX}{\mathord{\Box}}
\newcommand*{\DIA}{\mathord{\Diamond}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Nested sequents
\newcommand*\mdelim[3]{%
	\mathopen{}\left#1%
	#3%
	\right#2\mathclose{}%
}

\makeatletter
%\newcommand*\BR[2][]{\mdelim{\lbrack\strut^{#1}}{\rbrack}{#2}}
\newcommand*{\BR}{%
\@ifnextchar\i{\br@two}{%
\@ifnextchar\bgroup{\br@one}{% 
}}}
\newcommand*{\br@one}[1]{%
\def\br@{#1}%
\mdelim{\lbrack}{\rbrack}{\ifx\br@\empty\mkern 3mu\else #1\fi}%
}
\newcommand*{\br@two}[3]{%
\def\br@{#3}%
\mdelim{\lbrack\strut^{#2}}{\rbrack}{\ifx\br@\empty\mkern 3mu\else #3\fi}%
}
%
\newcommand*{\bBR}{%
\@ifnextchar\i{\bbr@two}{%
\@ifnextchar\bgroup{\bbr@one}{% 
}}}
\newcommand*{\bbr@one}[1]{%
\def\br@{#1}%
\mdelim{\llbracket}{\rrbracket}{\ifx\br@\empty\mkern 3mu\else #1\fi}%
}
\newcommand*{\bbr@two}[3]{%
\def\br@{#3}%
\mdelim{\llbracket\strut^{#2}}{\rrbracket}{\ifx\br@\empty\mkern 3mu\else #3\fi}%
}
%
%\newcommand*{\@makeoperator}[2]{
%	\newcommand*{#1}{%
%		\mathrm{#2}\mdelim{\lparen}{\rparen}
%	}
%}
%%
%\@makeoperator{\fm}{fm^{\n}}
%\@makeoperator{\ofm}{fm^{\o}}
%\@makeoperator{\hfm}{fm^{\h}}
%%
%\@makeoperator{\depth}{dp}
%\@makeoperator{\rank}{rk}
%\@makeoperator{\height}{ht}
%\@makeoperator{\tree}{tree}
%\@makeoperator{\graph}{graph}
\makeatother
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% contexts
\makeatletter
\newcommand*{\cxs}{%
\@ifnextchar\i{\cxs@two}{%
\@ifnextchar\bgroup{\cxs@one}{% % \bgroup is the same as {
}}}
\newcommand*{\cxs@one}[1]{%
\def\cxs@{#1}%
\mdelim{\lbrace}{\rbrace}{\ifx\cxs@\empty\mkern 3mu\else #1\fi}%
\cxs@one@decor%
}
\newcommand*{\cxs@two}[3]{%
\def\cxs@{#3}%
\mdelim{\lbrace\strut^{#2}}{\rbrace}{\ifx\cxs@\empty\mkern 3mu\else #3\fi}%
\cxs@one@decor%
}
\def\cxs@one@decor{%
\@ifnextchar_{\cxs@one@sub}{%
\@ifnextchar^{\cxs@one@sup}{%
	\@ifnextchar\dots{\@firstoftwo{\dotsm\cxs@one@decor}}{%
		\@ifnextchar[{\cxs@one@arg}%]
		\cxs}}}%
}
\def\cxs@one@sub_#1{_{#1}\cxs@one@decor}
\def\cxs@one@sup^#1{^{#1}\cxs@one@decor}
\def\cxs@one@arg[#1]{{#1}\cxs@one@decor}

%% Change these as needed, but keep the \cx@continuation at the end
\def\cx@delete@always#1{{#1}^{\ast}\cx@continuation}
\def\cx@delete@right#1*{{#1}^{\star}\cx@continuation}
\def\cx@delete@focus#1\f{{#1}^{\scriptscriptstyle\not{\langle}{\rangle}}\cx@continuation}

\def\cx@delete@star#1*{%
\@ifnextchar*{\cx@delete@right{#1}}{\cx@delete@always{#1}}%
}

\newcommand*{\@makecontextual}[2]{
\newcommand*{#1}{%
\@ifnextchar*{\cx@delete@star{#2}}{%
	\@ifnextchar\f{\cx@delete@focus{#2}}{%
		#2\cx@continuation}}%
}%
}
\newcommand*{\cx@continuation}[1][]{_{#1}\cxs}

\@makecontextual{\Ex}{}

\@makecontextual{\Cx}{\Gamma}
\@makecontextual{\Dx}{\Delta}
\@makecontextual{\Lx}{\Lambda}
\@makecontextual{\Rx}{\Pi}

\@makecontextual{\LLx}{\Omega}
\@makecontextual{\RRx}{\Xi}
\@makecontextual{\CCx}{\Sigma}
\@makecontextual{\DDx}{\Theta}
\makeatother
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% intuitionistic
\newcommand*{\rt}[1]{#1^\circ}
\newcommand*{\lf}[1]{#1^\bullet}

%%% Extracting symbols from MnSymbol 
\DeclareFontFamily{U} {MnSymbolC}{}
%
\DeclareFontShape{U}{MnSymbolC}{m}{n}{
<-6>  MnSymbolC5
<6-7>  MnSymbolC6
<7-8>  MnSymbolC7
<8-9>  MnSymbolC8
<9-10> MnSymbolC9
<10-12> MnSymbolC10
<12->   MnSymbolC12}{}
\DeclareFontShape{U}{MnSymbolC}{b}{n}{
<-6>  MnSymbolC-Bold5
<6-7>  MnSymbolC-Bold6
<7-8>  MnSymbolC-Bold7
<8-9>  MnSymbolC-Bold8
<9-10> MnSymbolC-Bold9
<10-12> MnSymbolC-Bold10
<12->   MnSymbolC-Bold12}{}
%
\DeclareSymbolFont{MnSyC}         {U}  {MnSymbolC}{m}{n}
%
\DeclareMathSymbol{\diamondplus}{\mathbin}{MnSyC}{124}
\DeclareMathSymbol{\boxtimes}{\mathbin}{MnSyC}{117}

%%% Labelled sequents
\newcommand{\SEQ}{\Rightarrow}
\newcommand*{\DD}{\mathcal{D}}
%\newcommand*{\rn}[1]  {\ensuremath{\mathsf{#1}}}
%\newcommand*{\invr}[1]{#1^\bullet}
%\newcommand*{\rel}{R}

%%% Labelled rules
\newcommand*{\labrn}[2][]  {\rn{#2}_{#1}}%^{\lab}}}
\newcommand*{\rlabrn}[2][]  {\rn{#2}_\rn{R#1}}%^\lab}}
\newcommand*{\llabrn}[2][]  {\rn{#2}_\rn{L#1}}%^\lab}}
%
\newcommand*{\brsym}{\boxtimes}%\mathord{\scalebox{.8}{$\blacksquare$}}}
\newcommand*{\diasym}{\diamondplus}%\mathord{\blacklozenge}}
%
\newcommand*{\boxbrn}[1]{\brsym_\rn{#1}}%^{\lab}}}
\newcommand*{\diabrn}[1]{\diasym_\rn{#1}}

\newcommand*{\labIKp}{\lab\IK_{\le}}
\newcommand*{\nIKp}{\n\IK_{\le}}

@@

\section{Introduction}
Structural proof theoretic accounts of modal logic can adopt the paradigm of \emph{labelled deduction} in the form of e.g.~labelled sequent systems~\cite{vigano2000,negri2005}, or the one of \emph{unlabelled deduction} in the form of e.g.~nested sequent systems~\cite{brunnler2009deep,poggiolesi2009method}.

@@

Similarly, proof systems have been designed for intuitionistic modal logic both as labelled sequent systems~\cite{simpson1994} and as nested sequent systems~\cite{strassburger2013,kuznets:strassburger:maehara,galmiche2018}.
%
Surprisingly, in nested and labelled sequents, extending the sequent structure with the same \emph{one} extra element is enough to obtain sound and complete systems.
%

@@

Incorporating the preorder symbol into the syntax, however,  allowed us~\cite{marin:morales:strassburger:hal} to %also obtain a sound and complete proof system for the intuitionistic modal logic extended with axiom $\rn{4}_\rn\DIA$, by designing the following rule:
%$$\scalebox{.9}{$\vlinf{\diabrn{4}}{\text{\footnotesize $\lb{u'}$ fresh}}{\B, \accs wv, \accs vu, \Left \SEQ \Right}{\B, \accs wv, \accs vu, \futs{u}{u'}, \accs{w}{u'}, \Left \SEQ \Right}$}$$
%Consequently, we can easily 
translate the conditions in~\eqref{eq:4frame} into separate inference rules for $\ax{4_{\BOX}}$ and $\ax{4_{\DIA}}$:\vspace*{-.2cm}
\begin{equation*}\small
\label{eq:4rules}
\vlinf{\rn{4_{\BOX}}}{\text{\footnotesize$\lb{u}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
	\B, \accs xy, \accs yz, \accs {u}z, \futs x{u},\Left \SEQ \Right}
\quad
\vlinf{\rn{4_{\DIA}}}{\text{\footnotesize$\lb{u}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
	\B, \accs xy, \accs yz, \accs {x}{u}, \futs z{u},\Left \SEQ \Right}  
\end{equation*}

%This allows us to define cut-free deductive systems for a wide range of logics that could not be treated before.

@@

%The generalisation of these results to our fully structured framework is subject of ongoing study.

These extensions for $\labIKp$ are already proveṇ̣~\cite{marin:morales:strassburger:hal} but these results for our fully nested sequent system are subject of ongoing study.

%The generalisation of these results to our fully structured framework is subject of ongoing study.
